---
title: "pd_kinematics"
author: "Valentijn Prové"
date: "27-10-2021"
output: html_document
---

```{r load packages, include=FALSE}
library(rstudioapi)
library(readbulk)
library(signal) #for butterworth filter
library(readxl)
library(ggplot2)
library(plotly)
library(dplyr)
library(pracma) #for peak finding
library(kza) #for kz smoothing
library(gganimate)
library(gridExtra)
library(ggpubr)

#for statistics
library(lme4)
library(effects)
library(rstatix)
library(insight)
library(lattice) 
library(ggbeeswarm)
library(MuMIn)
```

```{r define functions, include=FALSE}

#append ELAN annotations to timeseries
load.event <- function(time_ms_rec, g_d, column)
{
  output <- character(length = length(time_ms_rec))
  output <- NA
  for(i in g_d[,1])
  {
    #print(i)
    output <- ifelse((time_ms_rec >= i & time_ms_rec <= g_d[,2][g_d[,1] == i]), as.character(g_d[,column][g_d[,1]==i]), output)
  }
  return(output) #provides a vector with the length of the time series with annotations
}

#compute speed
speedXY.it <- function(x, y, time_millisecond)
{
  speed <- c(0, sqrt( rowSums( cbind(diff(x)^2, diff(y)^2)) ))
  speed <<- speed/(c(0, diff(time_millisecond)/1000))
}

#apply butterworth filter for smoothing
butter.it <- function(x)
{bf <- butter(1, 1/33, type="low")
x <- as.numeric(signal::filter(bf, x))}

#apply butterworth filter with forward and reverse filtering
butter.filtfilt <- function(x)
{bf <- butter(1, 1/33, type="low")
x <- as.numeric(signal::filtfilt(bf, x))}

#kinematic feature extraction:
kin.get <- function(MT)
{
  MT <- data.frame(MT)
  #peaks are based on smoothed speed data
  peaks    <- findpeaks(as.vector(scale(MT$xy_speed)), minpeakdistance = 8, minpeakheight = 0.5, threshold=0.5)
  
  #compute total submovements from all keypoints
  submovements <- sum(c(nrow(peaks), na.rm=TRUE))

  #compute average gesture space used by the gesture
  range.x <- max(MT$x)-min(MT$x)
  range.y <- max(MT$y)-min(MT$y)
  gspace <- round(1000*(range.x*range.y), digits = 3)
  gspace.log <- round(log(gspace+1), digits = 3) #log scale gspace because it explodes at high values
  
  #compute amplitudes (one-dimensional) 
  ampl.y <- round(max(MT$y), digits =  3)
  ampl.x <- round(max(MT$x), digits = 3)
  
  #compute triangular amplitude
  ampl.tri <- round(max(MT$tri_peri), digits =  3)
  
  #compute trajectory length
  trajectory <- round(sum(c(0, sqrt( rowSums( cbind(diff(MT$x)^2, diff(MT$y)^2)) ))), digits =  3)
  
  #compute peak velocity, based on smoothed speed data
  peak.vel <- round(max(MT$xy_speed), digits =  3)
  
  #compute total duration
  duration.ms <- max(MT$time_ms)-min(MT$time_ms)
  
  #bind everything into a single output object containing all the relevant kinematic features
  features <- cbind(submovements,range.x, range.y,
                    gspace, gspace.log, ampl.x, ampl.y, ampl.tri,
                    trajectory, peak.vel, duration.ms)
  return(features)
}
```

```{r set directories, include=FALSE}
datafolder <- paste0(dirname(dirname(rstudioapi::getSourceEditorContext()$path)), "/foreigner_talk_gesture/data/") #set your working directory to the folder where you saved the datasets
tsfolder <- paste0(dirname(dirname(rstudioapi::getSourceEditorContext()$path)), "/foreigner_talk_gesture/data/ts/") #folder with the timeseries output from OpenPose
```

```{r import data, include=FALSE}
ts <- read_bulk(tsfolder) # timeseries data

#ELAN annotations
annot.master <- read.csv(paste0(datafolder, "annot_master.csv")) # gesture units
annot.rh <- read.csv(paste0(datafolder, "annot_rh.csv")) # right hand gesture phases
annot.lh <- read.csv(paste0(datafolder, "annot_lh.csv")) #left hand gesture phases

#make sure that the filenames are structured in the same way in all datasets
ts$File <- gsub(".mp4_body25.csv", "\\1", ts$File)
annot.master$File <- gsub("_segm.eaf", "\\1", annot.master$File)
annot.rh$File <- gsub("_segm.eaf", "\\1", annot.rh$File)
annot.lh$File <- gsub("_segm.eaf", "\\1", annot.lh$File)
```

Loop through the different video files to create a dataframe containing the gesture and phase rates and a dataframe with the preprocessed timeseries data
```{r gesture rate and ts preprocess loop, include=FALSE}
#create dataframes to store the output
rates <- data.frame()
ts.prepr <- data.frame()

#loop through the files
for (video in unique(ts$File)) {
  print(video)
  tssub <- subset(ts, File == video)
  mastersub <- subset(annot.master, File == video)
  rhsub <- subset(annot.rh, File == video)
  lhsub <- subset(annot.lh, File == video)
  
  #add frame number
  tssub$frame <- seq(nrow(tssub))
  #add time in msec
  tssub$time_ms <- round(seq(1000/(25), (nrow(tssub)*(1000/25)), by = 1000/25 ))
  
  #center wrists to Neck and scale to MidHip
  #Neck becomes 0 and MidHip becomes 1
  tssub$y7_bs <- -1*((tssub$y7-tssub$y1)/(tssub$y8-tssub$y1))
  tssub$y4_bs <- -1*((tssub$y4-tssub$y1)/(tssub$y8-tssub$y1))
  tssub$x7_bs <- (tssub$x7-tssub$x1)/(tssub$y8-tssub$y1)
  tssub$x4_bs_rev <- -1*((tssub$x4-tssub$x1)/(tssub$y8-tssub$y1)) #reverse x-axis of right hand for amplitude interpretation
  tssub$x4_bs <- (tssub$x4-tssub$x1)/(tssub$y8-tssub$y1)
  
  tssub$x1_bs <- (tssub$x1-tssub$x1)/(tssub$y8-tssub$y1)
  tssub$y1_bs <- -1*(tssub$y1-tssub$y1)/(tssub$y8-tssub$y1)
  tssub$x8_bs <- (tssub$x8-tssub$x1)/(tssub$y8-tssub$y1)
  tssub$y8_bs <- -1*(tssub$y8-tssub$y1)/(tssub$y8-tssub$y1)
  
  #NOTE on scaling: for a perfect normalization, the angular velocity should be calculated for the elbow (triangle with shoulder and wrist),
  #but we want absolute measures for gesture size
  
  #apply butterworth smoothing + forward and reverse filter
  tssub$x4_bf <- butter.filtfilt(tssub$x4_bs)
  tssub$y4_bf <- butter.filtfilt(tssub$y4_bs)
  tssub$x7_bf <- butter.filtfilt(tssub$x7_bs)
  tssub$y7_bf <- butter.filtfilt(tssub$y7_bs)
  
  #apply kz smoothing
  tssub$x4_kz <- kz(tssub$x4_bs, m = 5, k = 3)
  tssub$x4_kz_rev <- kz(tssub$x4_bs_rev, m = 5, k = 3)
  tssub$y4_kz <- kz(tssub$y4_bs, m = 5, k = 3)
  tssub$x7_kz <- kz(tssub$x7_bs, m = 5, k = 3)
  tssub$y7_kz <- kz(tssub$y7_bs, m = 5, k = 3)
  
  #compute triangle perimeter (sum of euclidian lengths of traingle between neck, hip and wrist key points)
  tssub$tri4_upper <- sqrt( rowSums( cbind((tssub$x4_kz)^2, (tssub$y4_kz)^2)) )#euclidian distance of upper side (neck to wrist)
  tssub$tri4_lower <- sqrt( rowSums( cbind((tssub$x4_kz)^2, (tssub$y4_kz+1)^2)) )#euclidian distance of lower side (hip to wrist)
  tssub$tri4_peri <- mapply(sum, 1, tssub$tri4_upper, tssub$tri4_lower, na.rm=TRUE) #distance between neck and hip is always 1 because of scaling procedure
  
  tssub$tri7_upper <- sqrt( rowSums( cbind((tssub$x7_kz)^2, (tssub$y7_kz)^2)) )#euclidian distance of upper side (neck to wrist)
  tssub$tri7_lower <- sqrt( rowSums( cbind((tssub$x7_kz)^2, (tssub$y7_kz+1)^2)) )#euclidian distance of lower side (hip to wrist)
  tssub$tri7_peri <- mapply(sum, 1, tssub$tri7_upper, tssub$tri7_lower, na.rm=TRUE) #distance between neck and hip is always 1 because of scaling procedure
  
  #compute velocity of left and right wrist (kz smooth)
  tssub$xy4_speed <- speedXY.it(tssub$x4_kz, tssub$y4_kz, tssub$time_ms)
  tssub$xy7_speed <- speedXY.it(tssub$x7_kz, tssub$y7_kz, tssub$time_ms)
  #apply bf or kz smoothing
  tssub$xy4_speed_bf <- c(0, butter.filtfilt(tssub$xy4_speed[is.finite(tssub$xy4_speed)]))
  tssub$xy7_speed_bf <- c(0, butter.filtfilt(tssub$xy7_speed[is.finite(tssub$xy7_speed)]))
  
  tssub$xy4_speed_kz <- c(0, kz(tssub$xy4_speed[is.finite(tssub$xy4_speed)], m = 5, k = 3))
  tssub$xy7_speed_kz <- c(0, kz(tssub$xy7_speed[is.finite(tssub$xy7_speed)], m = 5, k = 3))
  
  #load gesture annotations
  tssub$g_unit <- load.event(tssub$time_ms, mastersub, 4)
  tssub$phase_rh <- load.event(tssub$time_ms, rhsub, 4)
  tssub$phase_lh <- load.event(tssub$time_ms, lhsub, 4)
  
  #bind iteration
  ts.prepr <- rbind.data.frame(ts.prepr, tssub)
  
  #####
  
  #compute total description duration (=total file duration)
  descr.dur.sec <- round(max(tssub$time_ms), digits = 2)/1000
  descr.dur.ms <- round(max(tssub$time_ms), digits = 2)
  
  #remove non-gesture data
  tssub <- subset(tssub, !is.na(tssub$g_unit))
  
  #count gesture segments in file
  ges.freq <- length(unique(tssub$g_unit))
  
  #LOOP THROUGH ALL GESTURE SEGMENTS IN THIS FILE
  #each time a new file enters the loop, create new placeholder for gesture segments
  rates.ges <- data.frame()
  temp.dur <- vector()
  rates.phase <- data.frame()
  for (s in unique(tssub$g_unit)) {
    segmsub <- subset(tssub, g_unit == s) #subset of gesture unit
    
    ges.dur.ms <- max(segmsub$time_ms)-min(segmsub$time_ms) #compute duration of segment
    
    #MAKE NEW DATASET AND MERGE LEFT AND RIGHT WRISTS HORIZONTALLY FOR KINEMATIC FEATURES
    
    segmsub.4 <- subset(segmsub, !is.na(segmsub$g_unit), select = c(File, time_ms, g_unit, phase_rh)) #remove non-gesture data and create a dataset per hand

    names(segmsub.4)[names(segmsub.4) == 'phase_rh'] <- 'phase'
    if(nrow(segmsub.4) >= 1){
      segmsub.4$hand <- "rh" 
      }
    
    segmsub.4$phase <- gsub("RH", "\\1", segmsub.4$phase)
    
    segmsub.7 <- subset(segmsub, !is.na(segmsub$g_unit), select = c(File, time_ms, g_unit, phase_lh))           #remove non-gesture data and create a dataset per hand

    names(segmsub.7)[names(segmsub.7) == 'phase_lh'] <- 'phase'
    if(nrow(segmsub.7) >= 1){
      segmsub.7$hand <- "lh" 
      }
    
    segmsub.7$phase <- gsub("LH", "\\1", segmsub.7$phase)
    
    segmsub.4 <- subset(segmsub.4, phase != "rest")
    segmsub.7 <- subset(segmsub.7, phase != "rest")
    
    segmsub.4$phase[segmsub.4$phase == "part_retr"] <- "retr"
    segmsub.7$phase[segmsub.7$phase == "part_retr"] <- "retr"
    
    segmsub.4$g_unit.id <- paste0(segmsub.4$File, segmsub.4$g_unit, segmsub.4$hand)
    segmsub.7$g_unit.id <- paste0(segmsub.7$File,  segmsub.7$g_unit, segmsub.7$hand)
    
    segmsub.4$phase.id <- paste0(segmsub.4$File, segmsub.4$g_unit, segmsub.4$hand, segmsub.4$phase)
    segmsub.7$phase.id <- paste0(segmsub.7$File,  segmsub.7$g_unit, segmsub.7$hand, segmsub.7$phase)
    
    segmsub.bh <- rbind.data.frame(segmsub.4, segmsub.7)
    
    temp.hand <- data.frame()
    for (h in unique(segmsub.bh$g_unit.id)) {
      handsub <- subset(segmsub.bh, g_unit.id == h)
      
      #in this segment, assess for each phase whether it is present or not
      hold <- ifelse(length(grep("hold", handsub$phase) >= 1), 1, 0)
      stroke <- ifelse(length(grep("stroke", handsub$phase) >= 1), 1, 0)
      prep <- ifelse(length(grep("prep", handsub$phase) >= 1), 1, 0)
      retr <- ifelse(length(grep("retr", handsub$phase) >= 1), 1, 0)
      
      hand.feats <- cbind(stroke, hold, prep, retr, as.vector(handsub$g_unit.id[1]))
      temp.hand <- rbind.data.frame(temp.hand, hand.feats)
    }
    
    rates.ges <- rbind.data.frame(rates.ges, temp.hand)
    
    temp.dur <- c(temp.dur, ges.dur.ms)
    
  }
  
  #compute proportional gesture time
  PoT <- sum(temp.dur)/descr.dur.ms
  
  stroke.freq <- sum(as.numeric(rates.ges$stroke))
  hold.freq <- sum(as.numeric(rates.ges$hold))
  prep.freq <- sum(as.numeric(rates.ges$prep))
  retr.freq <- sum(as.numeric(rates.ges$retr))
  
  stroke.rate.min <- (stroke.freq/descr.dur.sec)*60
  hold.rate.min <- (hold.freq/descr.dur.sec)*60
  prep.rate.min <- (prep.freq/descr.dur.sec)*60
  retr.rate.min <- (retr.freq/descr.dur.sec)*60
  
  ges.rate.min <- (ges.freq/descr.dur.sec)*60
  
  feats.file <- cbind(ges.freq, descr.dur.sec, ges.rate.min, PoT, stroke.freq, hold.freq, prep.freq, retr.freq,
                      stroke.rate.min, hold.rate.min, prep.rate.min, retr.rate.min, as.vector(segmsub$File[1]))
  
  #accumulate computations per file
  rates <- rbind.data.frame(rates, feats.file)

  
  #bind iteration
  #ts.prepr <- rbind.data.frame(ts.prepr, tssub)
}

#add the column name for File
colnames(rates) <- c(colnames(rates[1:12]), "File")

#inspect the dataframes
head(ts.prepr)
head(rates)
```

We used two methods to check our preprocessing procedure. First, we created video plots that we synchronized with the original videos in ELAN. Second, we used graphs to tune our KZ filter (and reject the Butterworth filter).
```{r make animated plots to check estimations in ELAN, eval=FALSE, include=FALSE}
for (i in unique(ts.prepr$File)) {
  fs <- subset(ts.prepr, File == i) 
  fs.rh <- data.frame("time_ms" = fs$time_ms,
                      "frame" = fs$frame,
                            "x" = fs$x4_kz,
                            "y" = fs$y4_kz)
  
  fs.rh$hand <- "rh"
  
  fs.lh <- data.frame("time_ms" = fs$time_ms,
                      "frame" = fs$frame,
                            "x" = fs$x7_kz,
                            "y" = fs$y7_kz)
  
  fs.lh$hand <- "lh"
  
  fs.binder <- rbind(fs.rh, fs.lh)
  
  p <- ggplot(
    fs.binder, 
    aes(x = x, y=y, colour = hand)) +
    geom_point(size = 0.5 )
  
  ggsave(p, file=paste0("plot_", i,".png"))
  
  p <- p + transition_time(time_ms) +
    transition_states(time_ms, transition_length = 0,
                      state_length = 1) +
    shadow_mark(future = T, size = 0.5)
  
  a <- animate(p, nframes = max(fs$frame), fps = 25, renderer = av_renderer())
  
  anim_save(paste0(f, ".mp4"), a)
  
}
```

```{r check x- and y-coordinates and velocity for a random file, eval=FALSE, include=FALSE}
sample <- subset(ts.prepr, File == "leuven5a_run1")

y.plot <- ggplot(
  sample, 
  aes(x = time_ms)) +
  geom_point(aes(y = y1, color="steelblue"), size = 0.5)+
  geom_point(aes(y = y4, color="darkred"), size = 0.5) +
  geom_point(aes(y = y7, color="orange"), size = 0.5) +
  geom_point(aes(y = y8, color="black"), size = 0.5) +
  ylab("y (pixels)") +
  scale_y_reverse() +
  ggtitle("y-positions of left and right wrist") +
  scale_color_identity(name = "Body part",
                       labels = c("hip", "right hand", "left hand", "neck"),
                       guide = "legend")

y.plot.bs <- ggplot(
  sample, 
  aes(x = time_ms)) +
  geom_point(aes(y = y1_bs, color="steelblue"), size = 0.5)+
  geom_point(aes(y = y4_bs, color="darkred"), size = 0.5) +
  geom_point(aes(y = y7_bs, color="orange"), size = 0.5) +
  geom_point(aes(y = y8_bs, color="black"), size = 0.5) +
  ylab("y (scaling score)") +
  theme(legend.position = "none") +
  scale_color_identity(name = "Body part",
                       labels = c("hip", "right hand", "left hand", "neck"),
                       guide = "legend")

y.plot.kz <- ggplot(
  sample, 
  aes(x = time_ms)) +
  geom_point(aes(y = y1_bs, color="steelblue"), size = 0.5)+
  geom_point(aes(y = y4_kz, color="darkred"), size = 0.5) +
  geom_point(aes(y = y7_kz, color="orange"), size = 0.5) +
  geom_point(aes(y = y8_bs, color="black"), size = 0.5) +
  ylab("y (KZ filter)") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text=element_text(size=16)) +
  scale_color_identity(name = "Body part",
                       labels = c("hip", "right hand", "left hand", "neck"),
                       guide = "legend")

x.plot.kz <- ggplot(
  sample, 
  aes(x = time_ms)) +
  geom_point(aes(y = y1_bs, color="steelblue"), size = 0.5)+
  geom_point(aes(y = -1*(x4_kz), color="darkred"), size = 0.5) +
  geom_point(aes(y = x7_kz, color="orange"), size = 0.5) +
  ylab("y (KZ filter)") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text=element_text(size=16)) +
  scale_color_identity(name = "Body part",
                       labels = c("hip", "right hand", "left hand", "neck"),
                       guide = "legend")

subplot(ggplotly(y.plot.kz, dynamicTicks = TRUE), ggplotly(x.plot.kz, dynamicTicks = TRUE), nrows = 2, heights = c(0.3,0.3), shareX = T)

#ggarrange(y.plot, y.plot.bs, y.plot.kz, common.legend = T, legend = "bottom", nrow = 3)
#subplot(plotly(y.plot), plotly(y.plot.bs), plotly(y.plot.kz), common.legend = T, legend = "bottom", nrow=3)
```

```{r delete ts data that is not correct, include=FALSE}
ts.prepr <- subset(ts.prepr, File != "antwerp2a_run1")
ts.prepr <- subset(ts.prepr, File != "antwerp2a_run2")
ts.prepr <- subset(ts.prepr, File != "leuven11a_run1")
ts.prepr <- subset(ts.prepr, File != "leuven11a_run2")
ts.prepr <- subset(ts.prepr, File != "luik5a_run1")
ts.prepr <- subset(ts.prepr, File != "luik5a_run2")

ts.prepr <- subset(ts.prepr, File != "luik1a_run1")
ts.prepr <- subset(ts.prepr, File != "luik2b_run2")
ts.prepr <- subset(ts.prepr, File != "antwerp1a_run1")

rates <- subset(rates, File != "luik5a_run1")
```

```{r get kinematic measures, include=FALSE}
#make separate datasets for right and left hand

g.rh <- data.frame("time_ms" = ts.prepr$time_ms,
                       "x" = ts.prepr$x4_kz_rev,
                       "y" = ts.prepr$y4_kz,
                       "xy_speed" = ts.prepr$xy4_speed_kz,
                       "tri_peri" = ts.prepr$tri4_peri,
                       "File" = ts.prepr$File,
                       "g_unit" = ts.prepr$g_unit,
                       "phase" = ts.prepr$phase_rh, "hand" = "rh")

g.rh <- subset(g.rh, !is.na(phase)) #delete ts data where phase is NA (no gesture)
g.rh <- subset(g.rh, phase != "rest") #delete ts data where phase is rest

g.lh <- data.frame("time_ms" = ts.prepr$time_ms,
                   "x" = ts.prepr$x7_kz,
                   "y" = ts.prepr$y7_kz,
                   "xy_speed" = ts.prepr$xy7_speed_kz,
                   "tri_peri" = ts.prepr$tri7_peri,
                   "File" = ts.prepr$File,
                   "g_unit" = ts.prepr$g_unit,
                   "phase" = ts.prepr$phase_lh, "hand" = "lh")

g.lh <- subset(g.lh, !is.na(phase)) #delete ts data where phase is NA (no gesture)
g.lh <- subset(g.lh, phase != "rest") #delete ts data where phase is rest

g.rh$feats_id <- paste0(g.rh$File, g.rh$g_unit)
g.lh$feats_id <- paste0(g.lh$File, g.lh$g_unit)

#compute kinematic features for each hand

f.rh <- data.frame() #make dataframe for right hand
for (i in unique(g.rh$feats_id)) {
  id <- subset(g.rh, feats_id == i)
  cc <- cbind(kin.get(id),as.vector(id$feats_id[1]), as.vector(id$hand[1]))
  f.rh <- rbind.data.frame(f.rh, cc)
}

colnames(f.rh) <- c(colnames(f.rh[1:11]), "feats_id", "hand")

f.lh <- data.frame() #make dataframe for left hand
for (i in unique(g.lh$feats_id)) {
  id <- subset(g.lh, feats_id == i)
  cc <- cbind(kin.get(id),as.vector(id$feats_id[1]), as.vector(id$hand[1]))
  f.lh <- rbind.data.frame(f.lh, cc)
}

colnames(f.lh) <- c(colnames(f.lh[1:11]), "feats_id", "hand")

#merge left and right hand dataframes

f <- merge(f.rh, f.lh, by = "feats_id", all = T, suffixes = c(".rh",".lh"))

write.csv(f, paste0(datafolder, "f.csv"))
f <- read.csv(paste0(datafolder, "f.csv"))

#create one measure for both hands:
  #determine wether a gesture is one-handed (rh, lh) or two-handed (rhlh)
f$hand.rh[is.na(f$hand.rh)] <- ""
f$hand.lh[is.na(f$hand.lh)] <- ""
f$hand <- paste0(f$hand.rh, f$hand.lh)
  #calculate pairwise maximum so as to obtain a global measure for one gesture unit
f$gspace.log <- pmax(f$gspace.log.rh, f$gspace.log.lh, na.rm=TRUE)
f$ampl.x <- pmax(f$ampl.x.rh, f$ampl.x.lh, na.rm=TRUE)
f$ampl.y <- pmax(f$ampl.y.rh, f$ampl.y.lh, na.rm=TRUE)
f$ampl.tri <- pmax(f$ampl.tri.rh, f$ampl.tri.lh, na.rm=TRUE)
f$trajectory <- pmax(f$trajectory.rh, f$trajectory.lh, na.rm=TRUE)
f$peak.vel <- pmax(f$peak.vel.rh, f$peak.vel.lh, na.rm=TRUE)
f$duration.ms <- pmax(f$duration.ms.rh, f$duration.ms.lh, na.rm=TRUE)

head(f)
```

```{r append file information, include=FALSE}
info <- read.csv(paste0(datafolder, "lal_info.csv"), sep = ";", header = T) #import data

f$File <- gsub("segm_[0-9]*", "\\1", f$feats_id)
ts.prepr <- merge(ts.prepr, info, by = "File", all.y = F)
f <- merge(f, info, by = "File", all.y = F)
rates <- merge(rates, info, by = "File", all.y = F)

f$run <- gsub("[a-z]*[0-9]*[a-b]_", "\\1", f$File)
rates$run <- gsub("[a-z]*[0-9]*[a-b]_", "\\1", rates$File)
f$run_first <- "no"
f$run_first[grep("a_run1", f$File)] <- "yes"

f$subject <- gsub("_run[1-2]", "\\1", f$File)
f$subject <- gsub("antwerp", "AN", f$subject)
f$subject <- gsub("leuven", "LE", f$subject)
f$subject <- gsub("luik", "LU", f$subject)

rates$subject <- gsub("_run[1-2]", "\\1", rates$File)
rates$subject <- gsub("antwerp", "AN", rates$subject)
rates$subject <- gsub("leuven", "LE", rates$subject)
rates$subject <- gsub("luik", "LU", rates$subject)

rates <- subset(rates, subject != "AN2a")
rates <- rates[-5, ]
```

Untill here, we need to run every chunk to be able to work with the data
```{r write dataframes and load them again to obtain numerical variables, include=FALSE}
write.csv(rates, "rates.csv")
write.csv(f, "f.csv")

rates <- read.csv("rates.csv")
f <- read.csv("f.csv")
```

```{r plot gesture trajectories and amplitudes, eval=FALSE, include=FALSE}
#sample <- subset(ts.prepr, File == "leuven5a_run1")
sample <- ts.prepr

ts.rh <- data.frame("x" = sample$x4_kz,
                    "y" = sample$y4_kz,
                    "condition" = sample$condition,
                    "phase" = sample$phase_rh)

ts.rh$hand <- "rh"

ts.lh <- data.frame("x" = sample$x7_kz,
                    "y" = sample$y7_kz,
                    "condition" = sample$condition,
                    "phase" = sample$phase_rh)

ts.lh$hand <- "lh"

ts.binder <- rbind(ts.rh, ts.lh)

f$p1.condition.num.x <- f$condition
f$p1.condition.num.x[grep("nns", f$p1.condition.num.x)] <- -0.4
f$p1.condition.num.x[grep("ns", f$p1.condition.num.x)] <- -0.6
f$p1.condition.num.x <- as.numeric(f$p1.condition.num.x)

f$p1.condition.num.y <- f$condition
f$p1.condition.num.y[grep("nns", f$p1.condition.num.y)] <- -0.9
f$p1.condition.num.y[grep("ns", f$p1.condition.num.y)] <- -1.0
f$p1.condition.num.y <- as.numeric(f$p1.condition.num.y)

f$p2.condition.num.x <- f$condition
f$p2.condition.num.x[grep("nns", f$p2.condition.num.x)] <- 0.4
f$p2.condition.num.x[grep("ns", f$p2.condition.num.x)] <- 0.6
f$p2.condition.num.x <- as.numeric(f$p2.condition.num.x)

f$p2.condition.num.y <- f$condition
f$p2.condition.num.y[grep("nns", f$p2.condition.num.y)] <- -0.9
f$p2.condition.num.y[grep("ns", f$p2.condition.num.y)] <- -1.0
f$p2.condition.num.y <- as.numeric(f$p2.condition.num.y)

p1 <- ggplot() +
  geom_point(data = ts.rh, aes(x = x, y = y), color = "grey", size = 0.5, alpha = 0.2) +
  geom_point(data = ts.lh, aes(x = x, y = y, color = condition), size = 0.7, alpha = 0.2) +
  geom_boxplot(data = f, aes(x = p1.condition.num.x, y = ampl.y.lh, fill=condition), width = 0.6) +
  geom_boxplot(data = f, aes(x = ampl.x.lh, y = p1.condition.num.y, fill=condition), width = 0.6) +
  ggtitle("Left hands") +
  ylab("y-position") +
  xlab("x-position") +
  theme_bw()+
  coord_fixed(ratio = 1);p1

p2 <- ggplot() +
  geom_point(data = ts.lh, aes(x = x, y = y), color = "grey", size = 0.5, alpha = 0.2) +
  geom_point(data = ts.rh, aes(x = x, y = y, color = condition), size = 0.7, alpha = 0.2) + # must include argument label "data"
  geom_boxplot(data = f, aes(x = p2.condition.num.x, y = ampl.y.rh, fill=condition), width = 0.6) +
  geom_boxplot(data = f, aes(x = -1*(ampl.x.rh), y = p2.condition.num.y, fill=condition), width = 0.6) +
  ggtitle("Right hands") +
  ylab("y-position") +
  xlab("x-position") +
  theme_bw()+
  coord_fixed(ratio = 1);p2

ggarrange(p2, p1, common.legend = T, legend = "right", nrow = 1)
```

```{r illustrate, eval=FALSE, include=FALSE}
library(png)
im <- readPNG("leuven5a_run1_65591.png")


fs <- subset(ts.prepr, File == "leuven5a_run1")
fsg <- subset(fs, g_unit == "segm_9")

fs.rh <- data.frame("time_ms" = fs$time_ms,
                    "x" = fs$x4_kz,
                    "y" = fs$y4_kz)

fs.rh$hand <- "right"

fs.lh <- data.frame("time_ms" = fs$time_ms,
                    "x" = fs$x7_kz,
                    "y" = fs$y7_kz)

fs.lh$hand <- "left"

fs.binder <- rbind(fs.rh, fs.lh)

p <- ggplot(
  fs.binder, 
  aes(x = x, y= y, colour = hand)) +
  geom_point(size = 0.5 )+ 
  coord_fixed(ratio =1) +
  ylab("y-position")+
  xlab("x-position")+
  theme_bw() +
  guides(color = guide_legend(override.aes = list(size = 3))); p

p2 <- p + geom_point(data = fsg, aes(x = x4_kz, y=y4_kz), color = "black") + geom_point(data = fsg, aes(x = x7_kz, y=y7_kz), color = "black"); p2

ggplot(data = fsg, aes(x = x4_kz, y=y4_kz)) +
  geom_point(color = "black") +
  geom_point(data = fsg, aes(x = x7_kz, y=y7_kz), color = "black") +
  geom_point(size = 0.5 )+ 
  coord_fixed(ratio =1) +
  ylab("Y-position")+
  xlab("X-position")+
  theme_bw()

ggsave("fig_measures_dpi600.png", units="in", width=4, height=6, dpi=600)

```

```{r define maximally distinct colors for subjects}
cols <- c("#f4aba0",
"#0257d8",
"#199e13",
"#9b30b7",
"#4b9500",
"#654fd6",
"#edc139",
"#ba82ff",
"#a89c00",
"#ef54d1",
"#00d28f",
"#a20090",
"#83d98a",
"#ff468e",
"#00c6ba",
"#bd001a",
"#0086ec",
"#eb9b00",
"#8991ff",
"#818d00",
"#ff8fec",
"#4d6000",
"#cfb1ff",
"#d47500",
"#007ac8",
"#ff7826",
"#008cb9",
"#ff9831",
"#514a8c",
"#776900",
"#3d5787",
"#ff8f63",
"#009771",
"#b1004f",
"#9ad3a7",
"#8f3900",
"#00684a",
"#ffadbf",
"#826237",
"#793e6a",
"#ffa781")
```

```{r gesture rate}

rates <- subset(rates, File != "antwerp2a_run2")
rates <- subset(rates, File != "antwerp1b_run1")
rates <- subset(rates, File != "antwerp1b_run2")
rates <- subset(rates, File != "luik3a_run1")

mean(rates$ges.freq)
sd(rates$ges.freq)
mean(rates$descr.dur.sec)

rates %>%
  group_by(language) %>%
  get_summary_stats(hold.rate.min, type = "mean_sd")

rates$PoT <- as.numeric(rates$PoT)

rates %>%
  group_by(condition) %>%
  get_summary_stats(PoT, type = "mean_sd")

summary(a <- aov(ges.rate.min ~ condition + gender + language, data = rates))

summary(a <- aov(hold.rate.min ~ condition + gender + language, data = rates))

summary(a <- aov(PoT ~ condition + gender + language, data = rates))

res.aov <- anova_test(
  data = rates, dv = PoT, wid =subject
  ,within = run
  ,between = c(condition, gender)
)
get_anova_table(res.aov)

res.aov <- anova_test(
  data = rates, dv = ges.rate.min, wid =subject
  ,within = run
  ,between = c(condition, gender)
)
get_anova_table(res.aov)

res.aov <- anova_test(
  data = rates, dv = hold.rate.min, wid =subject
  ,within = run
  ,between = c(condition, gender)
)
get_anova_table(res.aov)

ggplot(rates, aes(x=run, y=PoT, fill=language)) + 
  geom_boxplot()
```

```{r PoT (proportion of gesture time)}
mean(rates$PoT)
sd(rates$PoT)

rates %>%
  group_by(condition) %>%
  get_summary_stats(PoT, type = "mean_sd")

summary(a <- aov(PoT ~ condition, data = rates))

ggplot(rates, aes(x=condition, y=PoT, fill=item)) + 
  geom_boxplot()

ggplot(rates, aes(run, PoT, shape=item)) +
  geom_point(aes(color=item)) +
  geom_line(aes(group = subject)) +
  facet_grid(.~condition) +
  theme_bw()

ggplot(rates, aes(language, ges.rate.min)) +
  geom_boxplot() +
  #geom_point()
  geom_beeswarm(data = rates, aes(language, ges.rate.min, color=subject),
                dodge.width = 0.9, alpha = 0.8) +
  scale_color_manual(values = cols) +
  facet_grid(.~condition) +
  theme_bw() +
  theme(legend.position = "none") +
  stat_summary(fun.y = mean, color = "red", geom = "line", aes(group = 1)) + 
  stat_summary(fun.y=mean, geom="point", shape = 1, color="red", size = 4)

```

```{r gesture size, eval=FALSE, include=FALSE}
f <- read.csv("f.csv")
#make a boxplot with gesture space values per subject
ggplot(f, aes(x=gspace.log, y=subject, fill=condition)) + 
  geom_boxplot()

#reorder some factor levels for intuitive interpretation
f$condition <- as.factor(f$condition)
levels(f$condition)
f$condition <- factor(f$condition,levels(f$condition)[c(2,1)])

#linear mixed effects model
  #gesture space
m0 = lmer(gspace.log ~ (1|subject) + (1|item), data=f, REML = F)
m1 = update(m0,.~. + condition)
m2 = update(m1,.~. + gender)
m3 = update(m2,.~. + language)

r.squaredGLMM(m1)

summary(m1)

fixed <- do.call(rbind.data.frame, as.data.frame(allEffects(m1)))

fixed$gspace.log <- fixed$fit

#FIGURE 5
ggplot(data = fixed, aes(x = condition, y = gspace.log)) +
  geom_beeswarm(data = f, aes(condition, gspace.log, color=subject), dodge.width = 0.9, alpha = 0.8) +
  #scale_color_manual(values=cols) +
  scale_color_manual(values = sample(gray.colors(41))) +
  geom_point(aes(group = 1), size = 3) +
  geom_line(aes(group = 1), color="black",size=1) + 
  geom_errorbar(aes(x = condition, ymin = lower, ymax = upper),
                color="red",width=0.1,size=1) +
  theme_bw() +
  theme(legend.position = "none",
        text = element_text(size=14),
        axis.text = element_text(size=12)) +
  ylab("Gesture space") +
  xlab("Condition") +
  geom_hline(yintercept = 4.335, linetype = "dashed", size = 1)+
  scale_x_discrete(labels=c("ns" = "NS", "nns" = "NNS"))

ggsave("fig_lmer_gspace_1200dpi.png", units="in", width=5, height=4, dpi=1200)

```

```{r brol?, eval=FALSE, include=FALSE}
#linear mixed effects model
  #horizontal amplitude

f$condition <- as.factor(f$condition)
levels(f$condition)
f$condition <- factor(f$condition,levels(f$condition)[c(2,1)])

m0 = lmer(ampl.y ~ (1|subject) + (1|item), data=f, REML = F)
m1 = update(m0,.~. + condition)
m2 = update(m1,.~. + gender)
m3 = update(m2,.~. + language)

summary(m1)
anova(m0, m1)

randoms <- ranef(m1, condVar=TRUE)
dotplot(randoms)

str(dd <- as.data.frame(randoms))
dd$condition <- "NS"
dd$condition[grep("LE", dd$grp)] <- "NNS"
dd <- subset(dd, grpvar != "item")

dd %>%
  arrange(condsd) %>%
  mutate(grp=factor(grp, levels=grp)) %>%
  ggplot(aes(y=grp,x= condval, color = condition)) +
  geom_point() +
  geom_errorbarh(aes(xmin=condval -2*condsd, xmax=condval +2*condsd), height=0) +
  theme_bw() +
  scale_x_continuous(n.breaks = 10)

test %>%
  arrange(range.y) %>%
  mutate(subject=factor(subject, levels=subject)) %>%
  ggplot(aes(y=subject,x= range.y, color = condition)) +
  geom_point() +
  #geom_errorbarh(aes(xmin=condval -2*condsd, xmax=condval +2*condsd), height=0) +
  theme_bw() +
  scale_x_continuous(n.breaks = 10)

r.squaredGLMM(m1)

fixed <- do.call(rbind.data.frame, as.data.frame(allEffects(m1)))

fixed$ampl.y <- fixed$fit

test <- data.frame()
for (i in unique(f$subject)) {
  subj <- subset(f, subject == i)
  subj$range.y <- max(subj$ampl.y) - min(subj$ampl.y)
  subj$range.x <- max(subj$ampl.x) - min(subj$ampl.x)
  subj$max.y <- max(subj$ampl.y)
  subj$max.x <- max(subj$ampl.x)
  subj$sd.y <- sd(subj$ampl.y)
  subj$sd.x <- sd(subj$ampl.x)
  test <- rbind.data.frame(test, subj)
}

test$neck <- "above"
test$neck[test$max.y < 0.05] <- "under"

t <- table(test$neck, test$condition)
chisq <- chisq.test(t)
chisq$residuals

test <- distinct(test, subject, .keep_all = T)

test %>%
  group_by(condition) %>%
  get_summary_stats(range.y, type = "mean_sd")

test$condition <- as.factor(test$condition)
levels(test$condition)
t.test(test$range.y ~ test$condition, var.equal = T)
boxplot(test$range.y ~ test$condition)

#test$subject <- as.factor(test$subject)

test %>%
  ggplot(aes(x = reorder(subject, range.y), y = ampl.y, shape = neck,
             color = condition)) +
  geom_point() +
  scale_color_manual(values=c("black","grey"),
                     name = "Condition", labels = c("NNS", "NS")) +
  scale_shape_manual(values=c(15,1),
                     name = "Highest amplitude", labels = c("Above neck", "Under neck")) +
  geom_hline(yintercept = 0.05, linetype = "solid", size = 1, color = "red")+
  annotate("text", x = 3, y = 0.1, label = "Neck", color = "red") +
  ylab("Vertical amplitude") +
  xlab("Gestures per subject") +
  theme_bw() +
  theme(axis.text.x=element_blank(),
        text = element_text(size=14),
        axis.text = element_text(size=12))

ggsave("fig_viz_amply_1200dpi.png", units="in", width=6, height=4, dpi=1200)
```

```{r brol2?, eval=FALSE, include=FALSE}

x0 = lmer(ampl.x ~ (1|subject) + (1|item), data=f, REML = F)
x1 = update(x0,.~. + condition)

y0 = lmer(ampl.y ~ (1|subject) + (1|item), data=f, REML = F)
y1 = update(y0,.~. + condition)

fixedx <- do.call(rbind.data.frame, as.data.frame(allEffects(x1)))
fixedy <- do.call(rbind.data.frame, as.data.frame(allEffects(y1)))

fixedx$ampl.x <- fixedx$fit
fixedy$ampl.y <- fixedy$fit

fpx <- ggplot(data = fixedx, aes(x = condition, y = ampl.x)) +
  geom_beeswarm(data = f, aes(condition, ampl.x, color=subject), dodge.width = 0.9, alpha = 0.8) +
  scale_color_manual(values=cols) +
  geom_point(aes(group = 1), size = 3) +
  geom_line(aes(group = 1), color="black",size=1) + 
  geom_errorbar(aes(x = condition, ymin = lower, ymax = upper),
                color="red",width=0.1,size=1) +
  theme_bw() +
  theme(legend.position = "none"); fpx

fpy <- ggplot(data = fixedy, aes(x = condition, y = ampl.y)) +
  geom_beeswarm(data = f, aes(condition, ampl.y, color=subject), dodge.width = 0.9, alpha = 0.8) +
  scale_color_manual(values=cols) +
  geom_point(aes(group = 1), size = 3) +
  geom_line(aes(group = 1), color="black",size=1) + 
  geom_errorbar(aes(x = condition, ymin = lower, ymax = upper),
                color="red",width=0.1,size=1) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_y_continuous(n.breaks = 16)

ggarrange(fpx, fpy, nrow = 1)

px <- test %>%
  #arrange(range.y) %>%
  #mutate(subject=factor(unique(subject), levels=subject)) %>%
  ggplot(aes(x = reorder(subject, -range.x), y = ampl.x, color = condition)) +
  geom_point() +
  scale_shape_manual(values=c(17,6)) +
  scale_size_manual(values=c(2,4)) +
  theme_bw() +
  theme(axis.text.x=element_blank()) +
  labs(x = "subject (ordered by vertical range)",
       y = "horizontal amplitude") +
  scale_y_continuous(n.breaks = 10)
```

```{r rest, eval=FALSE, include=FALSE}


cols <- c("#f4aba0",
"#0257d8",
"#199e13",
"#9b30b7",
"#4b9500",
"#654fd6",
"#edc139",
"#ba82ff",
"#a89c00",
"#ef54d1",
"#00d28f",
"#a20090",
"#83d98a",
"#ff468e",
"#00c6ba",
"#bd001a",
"#0086ec",
"#eb9b00",
"#8991ff",
"#818d00",
"#ff8fec",
"#4d6000",
"#cfb1ff",
"#d47500",
"#007ac8",
"#ff7826",
"#008cb9",
"#ff9831",
"#514a8c",
"#776900",
"#3d5787",
"#ff8f63",
"#009771",
"#b1004f",
"#9ad3a7",
"#8f3900",
"#00684a",
"#ffadbf",
"#826237",
"#793e6a",
"#ffa781")

library(ggbeeswarm)
+
  geom_beeswarm(data = f, aes(condition, gspace.log, color=subject, dodge.width = 1, alpha = 1)) +
  facet_grid(.~item)

ggplot(data = f, aes(run, gspace.log, color=item)) +
  geom_boxplot(aes(group=run), width = 0.5) +
  geom_beeswarm(dodge.width = 0.5, alpha = 1) +
  #scale_color_manual(values=cols) +
  facet_grid(.~condition) +
  theme_bw() +
  geom_errorbar(data = fixed, aes(x = condition, ymin = lower, ymax = upper),
                color="purple",width=0.1,size=1)

anova(m3, m2)
summary(m1)
plot(allEffects(m1))

str(allEffects(m1))

  #x-amplitude
m0 = lmer(ampl.x ~ (1|subject), data=f, REML = F)
m1 = update(m0,.~. + hand)
m2 = update(m1,.~. + condition)
m3 = update(m2,.~. + run)
m4 = lmer(ampl.x ~  hand + condition *run + (1|subject), data=f, REML = F)

anova(m1, m2)
summary(m3)
plot(allEffects(m3))

  #y-amplitude
m0 = lmer(ampl.y ~ (1|subject), data=f, REML = F)
m1 = update(m0,.~. + condition)
m2 = update(m1,.~. + run)
m3 = update(m2,.~. + language)

anova(m2, m1)
summary(m2)
plot(allEffects(m2))
```

```{r randoms, eval=FALSE, include=FALSE}
coef(m1)
summary(m1)

randoms <- ranef(m1, condVar=TRUE)
dotplot(randoms)

str(dd <- as.data.frame(randoms))
dd$condition <- "NS"
dd$condition[grep("LE", dd$grp)] <- "NNS"
dd$condition[grep("item", dd$grpvar)] <- "item"

#FIGURE 4

ggplot(dd, aes(y=grp,x=condval)) +
  geom_point() +
  geom_errorbarh(aes(xmin=condval -2*condsd, xmax=condval +2*condsd), height=0) +
  theme_bw() +
  theme(legend.position = "none",
        axis.text=element_text(size=10),
        axis.title=element_text(size=14),
        axis.text.y = element_blank(),
        strip.text = element_text(size = 14)) +
  facet_grid(rows = vars(grpvar), scales = "free_y") +
  labs(y="Grouping variable", x = "Conditional mean") +
  facet_grid(rows = vars(grpvar), scales = "free_y")

ggsave("fig_ranef_gspace_1200dpi.png", units="in", width=5, height=4, dpi=1200)


ggplot(dd, aes(y=grp,x=condval)) +
  geom_point() +
  geom_errorbarh(aes(xmin=condval -2*condsd, xmax=condval +2*condsd), height=0) +
  theme_bw() +
  facet_grid(rows = vars(grpvar), scales = "free_y") +
  scale_x_continuous(n.breaks = 10)
```

```{r peak velocity, eval=FALSE, include=FALSE}

f$mean.vel <- f$trajectory/f$duration.ms

m0 = lmer(peak.vel ~ (1|subject) + (1|item), data=f, REML = F)
m1 = update(m0,.~. + condition)
m2 = update(m1,.~. + gender)

anova(m0, m1)
summary(m1)
plot(allEffects(m1))

m0 = lmer(trajectory ~ (1|subject), data=f, REML = F)
m1 = update(m0,.~. + condition)
m2 = update(m1,.~. + language)

anova(m2, m1)
summary(m1)
plot(allEffects(m1))

fixedtraj <- do.call(rbind.data.frame, as.data.frame(allEffects(m1)))

m0 = lmer(duration.ms ~ (1|subject) + (1|item), data=f, REML = F)
m1 = update(m0,.~. + condition)
m2 = update(m1,.~. + gender)

anova(m0, m1)
summary(m1)
plot(allEffects(m1))

fixeddur<- do.call(rbind.data.frame, as.data.frame(allEffects(m1)))

f %>%
  group_by(condition) %>%
  get_summary_stats(c(trajectory, duration.ms, peak.vel), type = "mean_sd")
```

```{r hold duration, eval=FALSE, include=FALSE}
annot.rh$hand <- "rh"
annot.lh$hand <- "lh"
names(annot.rh)[names(annot.rh) == 'RH_phase'] <- 'phase'
names(annot.lh)[names(annot.lh) == 'LH_phase'] <- 'phase'
holds <- subset((rbind(annot.rh, annot.lh)), phase == "stroke")
holds$File <- gsub("_segm.eaf", "\\1", holds$File)

holds <- merge(holds, info, by = "File", all.y = F)

holds$run <- gsub("[a-z]*[0-9]*[a-b]_", "\\1", holds$File)

holds$subject <- gsub("_run[1-2]", "\\1", holds$File)
holds$subject <- gsub("antwerp", "AN", holds$subject)
holds$subject <- gsub("leuven", "LE", holds$subject)
holds$subject <- gsub("luik", "LU", holds$subject)

m0 = lmer(Duration...msec ~ (1|subject) + (1|item), data=holds, REML = F)
m1 = update(m0,.~. + condition)
m2 = update(m1,.~. + gender)
m3 = update(m2,.~. + language)

anova(m2, m3)
summary(m3)
plot(allEffects(m3))
```


